---
title: "Report"
author: "Tuva Norderud Jensen, Eirik Røys, Kathrine Brun"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyr)
library(here)
library(tidyverse)

```

## Original data

The original data was taken from the [medicaldata webpage](https://higgi13425.github.io/medicaldata/reference/licorice_gargle.html) and transformed into a not tidy format before it was given to us to use as part of the RMED901 exam in fall 2024.

The data is taken from a study of patients undergoing elective thoracic surgery requiring a double-lumen endotracheal tube. 236 adult patients were enrolled. Data on Gender, physical status, BMI, age, Mallampati score, smoking status, preoperative pain, surgery size, intervention and the outcomes (cough, sore throat and pain swallowing at various time points) are provided in the origional data. For more information on the data see [codebook](DATA/codebook.html).

## Reading data

```{r, include=FALSE}
data <- read_delim(here("DATA", "exam_dataset.txt"))

joindata <- read_delim(here("DATA", "exam_joindata.txt"))
```

```{r}
data

joindata
```

### Joining datasets

```{r}
combined_data <- data %>%
  full_join(joindata, join_by("patient_id"), relationship = "many-to-many")

combined_data
```

## Making Tidy Data

To tidy the data we had to:

-   Identify rows with any NA values
    -   We excluded the two patients that had NA values since they were missing all postOp values.
-   Change variable types
-   Rearrange the dataset
    -   Rename columns
    -   Combine month and year into one column
    -   Split preOp_ASA_Mallampati into two columns
    -   Add postop_cough_change_extubation, postop_throatpain_change and BMI_quartile column
    -   Change column coding gender to "Male" and "Female" instead of "0"/"1"
-   Remove duplicated columns and rows
-   Rearranged the dataset

#### Dealing with missing values

```{r identifying rows with NA}
na_rows <- combined_data %>%
  filter(if_any(everything(), is.na))

print(na_rows)
```

```{r Removing NA}
combined_data <- combined_data %>%
  drop_na()

```

#### Changing variable types

```{r changing variable types, echo=FALSE}
combined_data <- combined_data %>%
  mutate(preOp_ASA_Mallampati = as.factor(preOp_ASA_Mallampati),
         preOp_smoking = as.factor(preOp_smoking),
         preOp_pain = as.factor(preOp_pain),
         treat = as.factor(treat),
         pacu30min_cough = as.factor(pacu30min_cough),
         pacu90min_cough = as.factor(pacu90min_cough),
         postOp4hour_cough = as.factor(postOp4hour_cough),
         pod1am_cough = as.factor(pod1am_cough),
         extubation_cough = as.factor(extubation_cough))

str(combined_data)
```

#### Rearranging dataset

```{r rearranging dataset, echo=FALSE}
combined_data <- combined_data %>%
  # Renaming columns
  rename(gender = "1gender") %>%
  rename(BMI = `BMI kg/m2`) %>%
  rename(age = preOp_age) %>%
  rename(smoking = preOp_smoking) %>%
  # Combine month and year into a single column 'date' and convert to day-month-year format
  mutate(date = paste("01", month, year, sep = "-")) %>%
  mutate(date = dmy(date)) %>%
  # Split column 'preOp_ASA_Mallampati' into 'ASA_score' and 'Mallampati score'
  separate(preOp_ASA_Mallampati, into = c("ASA_score", "Mallampati_score"), sep = "_") %>%
  # Sort rows based on patient IDs
  arrange(patient_id) %>%
  #adding new coloumns to the dataset
  mutate(
    postop_cough_change_extubation = as.numeric(extubation_cough) - as.numeric(pod1am_cough),  
    postop_throatpain_change = as.numeric(pacu30min_throatPain) - as.numeric(pod1am_throatPain)) %>% 
  #cut BMI into quartiles (4 equal parts)
  mutate(BMI_quartile=cut(BMI,breaks = quantile (BMI, probs = seq(0, 1, by = 0.25), na.rm = TRUE), 
                          include.lowest = TRUE, 
                          labels = c("Q1", "Q2", "Q3", "Q4"))) %>% 
  
  # added column coding gender to "Male" and "Female" instead of "0"/"1"
  mutate(gender=factor(gender,levels=c(0,1),labels=c("Male", "Female"))) %>% 
  
  # Reordering columns and selecting which to keep
  select(patient_id, BMI, age, smoking, gender, date, everything(), -month, -year)
```

#### Removing duplicates

```{r Removing duplicate column}
unique(combined_data$preOp_gender == combined_data$gender)

combined_data$preOp_gender <- NULL
```

```{r Removing duplicate rows}
combined_data <- combined_data %>%
  group_by_all() %>%
  filter(n() == 1) %>%
  ungroup()

combined_data <- combined_data %>%
  distinct(.keep_all = TRUE)

combined_data %>%
  filter(patient_id==48)
```

#### Tidy data

```{r Create tidy data file, echo=FALSE}
write_csv2(combined_data, file = here("DATA", "clean_data.csv"))
```

## Stratifying data

Data was stratified by a gender and min, max, mean and sd of age was reported.

```{r Stratify gender}
combined_data %>% 
  filter(gender == "Female") %>% 
  summarise(min(age), max(age), mean(age), sd(age))
```

Data was also stratified by the following instructions:

-   BMI \<25
-   Gender = female
-   Age \> 50
-   extubation_cough = TRUE

```{r Stratify}
combined_data %>% 
  filter(BMI < 25 ) %>% 
  filter(gender == "Female") %>% 
  filter(age > 50) %>% 
  filter(extubation_cough = TRUE) %>% 
  summarise(min(age), max(age), mean(age), sd(age))
```

Table based on gender and BMI_quartile

```{r Gender and BMI_quartile table}
combined_data %>%
  count(gender, BMI_quartile) %>%
  arrange(gender, BMI_quartile)
```

## Plots

...

## Statistical analysis

...

at the end add packages etc. with sessionInfo()

```{r}
sessionInfo()
```
