---
title: "Report"
author: "Tuva Norderud Jensen, Eirik Røys, Kathrine Brun"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyr)
library(here)
library(tidyverse)

```

## Original data

The data comes from a randomized trial, [here are the details](DATA/codebook.html)

## Reading data

```{r, include=FALSE}
data <- read_delim(here("DATA", "exam_dataset.txt"))

joindata <- read_delim(here("DATA", "exam_joindata.txt"))
```

```{r}
data

joindata
```

### Joining datasets

```{r}
combined_data <- data %>%
  full_join(joindata, join_by("patient_id"), relationship = "many-to-many")

combined_data
```

## Making Tidy Data

To tidy the data we had to:

-   Identify rows with any NA values
    -   We excluded the two patients that had NA values since they were missing all postOp values.
-   Change variable types
-   Rearrange the dataset
    -   Rename columns
    -   Combine month and year into one column
    -   Split preOp_ASA_Mallampati into two columns
    -   Add postop_cough_change_extubation, postop_throatpain_change and BMI_quartile column
    -   Change column coding gender to "Male" and "Female" instead of "0"/"1"
-   Remove duplicated columns and rows
-   Rearranged the dataset

#### Dealing with missing values

```{r identifying rows with NA}
na_rows <- combined_data %>%
  filter(if_any(everything(), is.na))

print(na_rows)
```

```{r Removing NA}
combined_data <- combined_data %>%
  drop_na()

```

#### Changing variable types

```{r changing variable types, echo=FALSE}
combined_data <- combined_data %>%
  mutate(preOp_ASA_Mallampati = as.factor(preOp_ASA_Mallampati),
         preOp_smoking = as.factor(preOp_smoking),
         preOp_pain = as.factor(preOp_pain),
         treat = as.factor(treat),
         pacu30min_cough = as.factor(pacu30min_cough),
         pacu90min_cough = as.factor(pacu90min_cough),
         postOp4hour_cough = as.factor(postOp4hour_cough),
         pod1am_cough = as.factor(pod1am_cough),
         extubation_cough = as.factor(extubation_cough))

str(combined_data)
```

#### Rearranging dataset

```{r rearranging dataset, echo=FALSE}
combined_data <- combined_data %>%
  # Renaming columns
  rename(gender = "1gender") %>%
  rename(BMI = `BMI kg/m2`) %>%
  rename(age = preOp_age) %>%
  rename(smoking = preOp_smoking) %>%
  # Combine month and year into a single column 'date' and convert to day-month-year format
  mutate(date = paste("01", month, year, sep = "-")) %>%
  mutate(date = dmy(date)) %>%
  # Split column 'preOp_ASA_Mallampati' into 'ASA_score' and 'Mallampati score'
  separate(preOp_ASA_Mallampati, into = c("ASA_score", "Mallampati_score"), sep = "_") %>%
  # Sort rows based on patient IDs
  arrange(patient_id) %>%
  #adding new coloumns to the dataset
  mutate(
    postop_cough_change_extubation = as.numeric(extubation_cough) - as.numeric(pod1am_cough),  
    postop_throatpain_change = as.numeric(pacu30min_throatPain) - as.numeric(pod1am_throatPain)) %>% 
  #cut BMI into quartiles (4 equal parts)
  mutate(BMI_quartile=cut(BMI,breaks = quantile (BMI, probs = seq(0, 1, by = 0.25), na.rm = TRUE), 
                          include.lowest = TRUE, 
                          labels = c("Q1", "Q2", "Q3", "Q4"))) %>% 
  
  # added column coding gender to "Male" and "Female" instead of "0"/"1"
  mutate(gender=factor(gender,levels=c(0,1),labels=c("Male", "Female"))) %>% 
  
  # Reordering columns and selecting which to keep
  select(patient_id, BMI, age, smoking, gender, date, everything(), -month, -year)
```

#### Removing duplicates

```{r Removing duplicate column}
unique(combined_data$preOp_gender == combined_data$gender)

combined_data$preOp_gender <- NULL
```

```{r Removing duplicate rows}
combined_data <- combined_data %>%
  group_by_all() %>%
  filter(n() == 1) %>%
  ungroup()

combined_data <- combined_data %>%
  distinct(.keep_all = TRUE)

combined_data %>%
  filter(patient_id==48)
```

## Plots

...

## Statistical analysis

...

at the end add packages etc. with sessionInfo()

```{r}
sessionInfo()
```
