---
title: "Report"
author: "Tuva Norderud Jensen, Eirik Røys, Kathrine Brun"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(lubridate)
library(skimr)
library(ggcorrplot)
library(dplyr)
library(styler)
```

## Original data

The original data came from the [medicaldata webpage](https://higgi13425.github.io/medicaldata/reference/licorice_gargle.html) and transformed into a not tidy format before it was given to us to use as part of the RMED901 exam in fall 2024.

The data is taken from a study of patients undergoing elective thoracic surgery requiring a double-lumen endotracheal tube. 236 adult patients were enrolled. The study investigates whether gargling with licorice before anesthesia can prevent postoperative sore throat and coughing in these patients. Data on Gender, physical status, BMI, age, Mallampati score, smoking status, preoperative pain, surgery size, intervention and the outcomes (cough, sore throat and pain swallowing at various time points) are provided in the original data. For more information on the data see [codebook](DATA/codebook.html).

## Reading data

```{r, include=FALSE}
data <- read_delim(here("DATA", "exam_dataset.txt"))

joindata <- read_delim(here("DATA", "exam_joindata.txt"))
```

```{r}
data

joindata
```

### Joining datasets

```{r}
combined_data <- data %>%
  full_join(joindata, join_by("patient_id"), relationship = "many-to-many")

combined_data
```

## Making Tidy Data

To tidy the data we had to:

-   Identify rows with any NA values
    -   We excluded the two patients that had NA values since they were missing all postOp values.
-   Change variable types
-   Rearrange the dataset
    -   Rename columns
    -   Combine month and year into one column
    -   Split preOp_ASA_Mallampati into two columns
    -   Add postop_cough_change_extubation, postop_throatpain_change and BMI_quartile column
    -   Change column coding gender to "Male" and "Female" instead of "0"/"1"
-   Remove duplicated columns and rows
-   Rearranged the dataset

#### Dealing with missing values

```{r identifying rows with NA}
na_rows <- combined_data %>%
  filter(if_any(everything(), is.na))

print(na_rows)
```

```{r Removing NA}
combined_data <- combined_data %>%
  drop_na()

```

#### Changing variable types

```{r changing variable types, echo=FALSE}
combined_data <- combined_data %>%
  mutate(preOp_ASA_Mallampati = as.factor(preOp_ASA_Mallampati),
         preOp_smoking = as.factor(preOp_smoking),
         preOp_pain = as.factor(preOp_pain),
         treat = as.factor(treat),
         pacu30min_cough = as.factor(pacu30min_cough),
         pacu90min_cough = as.factor(pacu90min_cough),
         postOp4hour_cough = as.factor(postOp4hour_cough),
         pod1am_cough = as.factor(pod1am_cough),
         extubation_cough = as.factor(extubation_cough))

str(combined_data)
```

#### Rearranging dataset

-   renaming columns

-   combined month and year into a single column "date in month-year (my) format

-   split the coloumn ASA_Mallampati into two separate variables

-   sorted rows based in patient ID

-   added new columns showing change in cough and throat pain

-   cutting BMI into quartiles

-   Reordering columns and selecting which to keep

```{r rearranging dataset, echo=FALSE}
combined_data <- combined_data %>%
  # Renaming columns
  rename(gender = "1gender") %>%
  rename(BMI = `BMI kg/m2`) %>%
  rename(age = preOp_age) %>%
  rename(smoking = preOp_smoking) %>%
  # Combine month and year into a single column 'date' and convert to day-month-year format
  mutate(date = paste("01", month, year, sep = "-")) %>%
  mutate(date = dmy(date)) %>%
  # Split column 'preOp_ASA_Mallampati' into 'ASA_score' and 'Mallamdata:image/png;base64,iVBORw0KGgoAAAANSUhEUgAAABoAAAAaCAYAAACpSkzOAAABH0lEQVR4XmNgGAVDCqxdvzVhzcbNm6mBK6uqVqObDwf///9nXLt+SzG6JlJxTW3tVpBZ6OajgP3797Os3bClFV0zsbiuoXEfyAx0c7GCRTt3cq9Zv7UX3RBCuL6x8TBIL7p5eMGqVbv512zYMgvdMFy4san1FEgPujlEgbVrd0gCg3EJuqHouLm59TxILbp+ksCqjdvV12zcshLdcBhuamm7BFKDro8ssGbjRr01G7aux7Ckte0qSA5dPUVg9frNDkDLNsGDq7X9GkgMXR1VwNoNm0MgPmm/DmKjy49wwLjIbzM1MLq5GIBxgc9mhm2eFGGwGYQA42wvDI2kYrAZhADjdEyNpGKwGYQA42Q3DI2kYrAZBEGfy2aq4FEw5AAASl8F3R9fQqAAAAAASUVORK5CYII=pati score'
  separate(preOp_ASA_Mallampati, into = c("ASA_score", "Mallampati_score"), sep = "_") %>%
  # Sort rows based on patient IDs
  arrange(patient_id) %>%
  #adding new coloumns to the dataset
  mutate(
    postop_cough_change_extubation = as.numeric(extubation_cough) - as.numeric(pod1am_cough),  
    postop_throatpain_change = as.numeric(pacu30min_throatPain) - as.numeric(pod1am_throatPain)) %>% 
  #cut BMI into quartiles (4 equal parts)
  mutate(BMI_quartile=cut(BMI,breaks = quantile (BMI, probs = seq(0, 1, by = 0.25), na.rm = TRUE), 
                          include.lowest = TRUE, 
                          labels = c("Q1", "Q2", "Q3", "Q4"))) %>% 
  
  # added column coding gender to "Male" and "Female" instead of "0"/"1"
  mutate(gender=factor(gender,levels=c(0,1),labels=c("Male", "Female"))) %>% 
  
  # Reordering columns and selecting which to keep
  select(patient_id, BMI, age, smoking, gender, date, everything(), -month, -year)

combined_data
```

#### Removing duplicates

```{r Removing duplicate column}
unique(combined_data$preOp_gender == combined_data$gender)

combined_data$preOp_gender <- NULL
```

```{r Removing duplicate rows}
combined_data <- combined_data %>%
  group_by_all() %>%
  filter(n() == 1) %>%
  ungroup()

combined_data <- combined_data %>%
  distinct(.keep_all = TRUE)

combined_data %>%
  filter(patient_id==48)
```

======= \#### Tidy data

```{r Create tidy data file, echo=FALSE}
write_csv2(combined_data, file = here("DATA", "clean_data.csv"))
```

## Stratifying data

Data was stratified by a gender and min, max, mean and sd of age was reported.

```{r Stratify gender}
combined_data %>% 
  filter(gender == "Female") %>% 
  summarise(min(age), max(age), mean(age), sd(age))
```

Data was also stratified by the following instructions:

-   BMI \<25
-   Gender = female
-   Age \> 50
-   extubation_cough = TRUE

```{r Stratify}
combined_data %>% 
  filter(BMI < 25 ) %>% 
  filter(gender == "Female") %>% 
  filter(age > 50) %>% 
  filter(extubation_cough = TRUE) %>% 
  summarise(min(age), max(age), mean(age), sd(age))
```

Table based on gender and BMI_quartile

```{r Gender and BMI_quartile table}
combined_data %>%
  count(gender, BMI_quartile) %>%
  arrange(gender, BMI_quartile)
```

## Plots

```{r, include=TRUE}
data <- read.csv2(here("DATA", "clean_data.csv"))
```

# **Analyis: Relationship between variables**

## **Are there any correlated measurements?**

```{r correlation-plot, fig.width=9, fig.height=9}
# Create correlation matrix excluding non-numeric and irrelevant columns
corr <- data %>%
  select(-gender, -date, -BMI_quartile, -patient_id) %>%
  cor()

# Create matrix of correlation p-values
p.mat <- data %>%
  select(-gender, -date, -BMI_quartile, -patient_id) %>%
  cor_pmat()

# Plot the correlation matrix with p-values
corr_plot <- ggcorrplot(corr,
                        method = "circle",
                        hc.order = TRUE,
                        title = "Correlation plot",
                        outline.color = "white",
                        p.mat = p.mat
)

# Display plot
corr_plot


```

### Conclusion: several measurements are correlated: e.g., `pacu30min_swallowPain` and `pacu30min_throatpain`

```{r specify variables, include=FALSE, warning= FALSE}
data <- data %>%
  mutate(
    ASA_score = as.factor(ASA_score),
    Mallampati_score = as.factor(Mallampati_score),
    smoking = as.factor(smoking),
    preOp_pain = as.factor(preOp_pain),
    treat = as.factor(treat),
    pacu30min_throatPain = as.factor(pacu30min_throatPain),
    pacu90min_throatPain = as.factor(pacu90min_throatPain),
    postOp4hour_throatPain = as.factor(postOp4hour_throatPain),
    pod1am_throatPain = as.factor(pod1am_throatPain),
    pacu30min_cough = as.factor(pacu30min_cough),
    pacu90min_cough = as.factor(pacu90min_cough),
    postOp4hour_cough = as.factor(postOp4hour_cough),
    pod1am_cough = as.factor(pod1am_cough),
    extubation_cough = as.factor(extubation_cough)
  )

```

## **Does the `age` distribution depend on `treat`?**

```{r Age vs. Treat, include = TRUE, warning= FALSE}
# Boxplot Age vs. Treat
boxplot_treat <- ggplot(data, aes(x = treat, y = age, fill = treat)) +
  geom_boxplot() +
  scale_fill_manual(values = c("0" = "#E1E1E1", "1" = "#649CB8")) +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), limits = c(0, 100)) +
  labs(title = "Box-plot") +
  theme_classic() +
  coord_flip()

# Histogram Age vs. Treat
histogram_treat <- ggplot(data, aes(y = age, fill = treat)) +
  geom_histogram(binwidth = 10, position = "dodge", color = "black") +
  scale_fill_manual(values = c("0" = "#E1E1E1", "1" = "#649CB8")) +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), limits = c(0, 100)) +
  labs(title = "Histogram") +
  theme_classic() +
  coord_flip()

# Combine plots
Age_vs_treat_plot <- (boxplot_treat / histogram_treat) +
  plot_annotation(
    title = "Age vs. Treatment",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  )

# Display the plot
Age_vs_treat_plot

```

### Conclusion: the distribution of `age` does not appear to depend on `treatment`.

## **Does the `age` distribution of the patients depend on their `gender`?**

```{r Age vs. Gender, include = TRUE, warning= FALSE}

# Boxplot: Age vs. Gender
boxplot_gender <- ggplot(data, aes(x = gender, y = age, fill = gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Female" = "#FBADA7", "Male" = "#64B8B8")) +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), limits = c(0, 100)) +
  labs(title = "Box-plot") +
  theme_classic() +
  coord_flip()

# Histogram: Age vs. Gender
histogram_gender <- ggplot(data, aes(y = age, fill = gender)) +
  geom_histogram(binwidth = 10, position = "dodge", color = "black") +
  scale_fill_manual(values = c("Female" = "#FBADA7", "Male" = "#64B8B8")) +
  scale_y_continuous(breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), limits = c(0, 100)) +
  labs(title = "Histogram") +
  theme_classic() +
  coord_flip()

# Combine plots
Age_vs_gender_plot <- (boxplot_gender / histogram_gender) +
  plot_annotation(
    title = "Age vs. Gender",
    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  )

# Display the plot
Age_vs_gender_plot

```

### Conclusion: The distribution of `age` does not appear to depend on `gender`.

## **Do `BMI` and `age` have a linear relationship?**

```{r Age vs. BMI, include = TRUE, warning= FALSE}

# ---- Do BMI and age have a linear relationship? -----------------------------------------

Regresssion_plot <- ggplot(data, aes(x = age, y = as.numeric(BMI))) +
  geom_point(alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", color = "steelblue", se = TRUE, size = 2, fill = "lightblue") +
  labs(
    title = "BMI and age relationship",
    x = " Age of Patients",
    y = "BMI"
  ) +
  theme_classic()

# Display the plot
Regresssion_plot

```

### Conclusion: it appears to be a linear relationship between `BMI` and `age`.

## **Does the `preOp_pain` change with the `age` of the patients?**

```{r preOppain vs. age, include = TRUE, warning= FALSE}

sum(data$preOp_pain == 1)
sum(data$preOp_pain != 1)

```

#### Conclusion: Only two patients (out of 249) reported `preOp_pain` - there is insufficient data to test for a relationship.

## Statistical analysis

#### Does the treatment depend on the preoperative smoking?

```{r treat smoking}
logistic_model <- glm(data$treat ~ data$smoking, family = binomial())
summary(logistic_model)
```

The likelihood of receiving licorice treatment do not depend on Preoperative smoking status (P \> 0.05).

#### Does the treatment depend on the gender of the patient?

```{r treat gender}
logistic_model_treatment_gender <- glm(data$treat ~ data$gender, family = binomial())
summary(logistic_model_treatment_gender)
```

The likelihood of receiving licorice treatment does not depend on gender (P \> 0.05).

#### Does the treatment depend on whether the patient had a preoperative pain?

```{r treat preop pain}
logistic_model_treatment_preoppain <- glm(data$treat ~ data$preOp_pain, family = binomial())
summary(logistic_model_treatment_preoppain)
```

The likelihood of receiving licorice treatment does not depend on preoperative pain (P \> 0.05).

#### Was the treatment with licorice gargle reducing the risk of post operative throat pain?

```{r risk of postop pain}
# Fit the logistic regression between 'treat' and throat pain reported at four time intervals after surgery (30min, 90min, 4hours and 1am)
model_pacu30min_throatPain <- glm(pacu30min_throatPain ~ treat, data = data, family = binomial)
model_pacu90min_throatPain <- glm(pacu90min_throatPain ~ treat, data = data, family = binomial)
model_postOp4hour_throatPain <- glm(postOp4hour_throatPain ~ treat, data = data, family = binomial)
model_pod1am_throatPain <- glm(pod1am_throatPain ~ treat, data = data, family = binomial)

# The summary from all four model show that licorice treatment significantly reduces throat pain after surgery at all time points (P < 0.05)
summary(model_pacu30min_throatPain)
summary(model_pacu90min_throatPain)
summary(model_postOp4hour_throatPain)
summary(model_pod1am_throatPain)

# To estimate the effect size we can also calculate the odds ratio from the logistic regression models (from the log-odds)
odds_ratio_pacu30min <- (exp(coef(model_pacu30min_throatPain)["treat1"]) - 1) * 100
odds_ratio_pacu90min <- (exp(coef(model_pacu90min_throatPain)["treat1"]) - 1) * 100
odds_ratio_postOp4hours <- (exp(coef(model_postOp4hour_throatPain)["treat1"]) - 1) * 100
odds_ratio_pod1am <- (exp(coef(model_pod1am_throatPain)["treat1"]) - 1) * 100

# Print the odds
odds_ratio_pacu30min
odds_ratio_pacu90min
odds_ratio_postOp4hours
odds_ratio_pod1am
```

From the odds ratios we see that licorice treatment reduces the risk of throat pain relative to those who did not receive the licorice treatment (but got sugar instead) by:

-   65.4% after 30 minutes

-   80.6% after 90 minutes,

-   70.3% after 4 hours

-   66.8% on the first postoperative morning

## Session info

```{r, echo=FALSE}
sessionInfo()
```
